Resource-ratio theory in metaecosystems

### 2013-11-05 ms_resource_ratio.tex
Edits on the intro

The interaction between community assembly (determining the species composition of local communities) and ecosystem attributes (determining the stocks and fluxes of nutrients and energy) is complex because it depends on so many distinct processes including nutrient consumption, recycling, export and import of materials by the biota.  The effects are even more complex because of stoichiometric relations that constrain how processes involving one element may be connected to another (e.g. nitrogen and phosphorous fluxes through an ecosystem is strongly linked in some processes such as plant uptake and not others such as recycling). Recent work on these issues has revealed important consequences of such stoichiometric mechanisms that operate within ecosystems (Sterner and Elser, Loreau).  However exchanges of materials and organisms between ecosystems may also be important in more spatially extended ‘meta-ecosystem’s (Loreau et al, Gravel et al 2010a, Gravel et al. 2010b) although their effects on stoichiometric mechanisms have not yet been well studied (but see Ryabov and Blasius 2011).

In the limit as exchanges between two ecosystems increase via enhanced diffusion there should be a pronounced homogenizing effects since the dynamics of the combined ecosystems will be identical to each subsystem if they are completely mixed.  However the effects at lower exchange rates are harder to predict since it may depend on which types of materials or organisms are the ones that move between ecosystems.  For example, if the stoichiometry of dispersing consumers, of dissolved available nutrients or of detritus are distinct, the stoichiometry of exchanges between ecosystems may be affected  in significantly different ways.

Here we use ‘niche based’ (Leibold 1995) approaches to resource competition between two consumers growing on two essential nutrient resources (Tilman 1982) in a metaecosystem consisting of two patches in source-sink situations (Gravel et al. 2010a).  We account for internal resource cycling within each patch (Daufresne and Hedin) and ask how the diffusive (unbiased, dependent on density or concentration differences between patches) exchanges of dissolved nutrients, detritus or consumers may differ in the consequences for consumer community composition and coexistence and for the concentrations of available nutrients.  We find that each type of exchange affects the process in fundamentally distinct ways. Diffusion affect the stoichiometry of the exchanges between patches due to variability in the stoichiometry of each type of material and on the indirect effects they have on resource competition.   Exchanges of detritus (assuming consumers and their detritus has a fixed stoichiometry) homogenize the productivity of the two patches without affecting any local stoichiometric effects, that exchanges of dissolved nutrients will homogenize both the productivity and stoichiometry of the two patches, and that exchanges of live consumers will have effects that maintain stoichiometry but affect nutrient depletion in patches.  We also show that the joint effects of detritus exchanges can affect coexistence and may generate alternate stable states in local communities.

>>>>

The interaction between community assembly and ecosystem attributes depends on many distinct processes including nutrient consumption, recycling, export and import of materials by the biota.  It is even more complex because of stoichiometric relationships constraining how processes involving one element may be connected to another. For instance, nitrogen and phosphorous fluxes through an ecosystem are strongly linked through various processes such as plant uptake, and not others such as mineralization. Recent work on these issues has revealed important consequences of such stoichiometric mechanisms that operate within ecosystems (Sterner and Elser, Loreau).  However exchanges of materials and organisms between ecosystems may also be important in more spatially extended ‘meta-ecosystem’s (Loreau et al, Gravel et al 2010a, Gravel et al. 2010b) although their effects on stoichiometric mechanisms have not yet been well studied (but see Ryabov and Blasius 2011).

The resource-ratio theory of plant coexistence (Tilman, 1982; 1988) has been instrumental in our understanding of the intimate linkage between stochiometry, community assembly and ecosystem functioning. The theory applies to two resources the R* principle of competition theory. Its main prediction is that stable coexistence between two species requires a particular ratio of the two limiting nutrients. Owing to its accessible graphical representation, the theory has a central position in most ecological textbooks (e.g. Begon et al. 2006). The theory was also further developped to derive of a vast array of secondary predictions such as the impat of resource heterogeneity and of fertilization on species richness and successional dynamics (Tilman 1982, 1985). The theory has also been updtated several times since its formulation. Notably, Leibold (1995) extended the theory to include the effect of herbivory on species coexistence, and Chase et Leibold (2002) built on the graphical representation to integrate the concepts of Eltonian and Grinellian niches. Daufresne and Hedin (2005) also recasted the theory with the consumer-resource mathematical formalism and included the effect of nutrient recycling on plant coexistence. While the theory is implicitly spatial in its original formulation and subsequent applications, there are surprisingly only few attemps at integrating with metacommunity ecology (see for instance Mouquet et al. 2006).

The resource ratio theory builds on the idea of spatial heterogeneity in the ratio of limiting resources to predict highly diverse communities (Tilman, 1982). This prediction does apply to various spatial scales, from the individual to individual variation in soil properties, to landscape variations. The theory does not however consider the impact of spatial exchanges of plants, nutrients and other materials between localities. Both metacommunity (Mouquet and Loreau, 2002; Abrams et al. 2004) and metaecosystem (Gravel et al. 2010) theory in source-sinks settings have shown that the outcome of competitive interactions could be significantly  altered by these flows. In the limit as exchanges between two ecosystems increase, there should be a pronounced homogenization leading to the dominance of one or a few strong competitors. The effects of spatial exchanges at lower rates are however harder to predict since it may depend on which types of materials or organisms are moving between ecosystems.  For example, if the stoichiometry of dispersing consumers, of dissolved available nutrients or of detritus are distinct, the stoichiometry of exchanges between ecosystems may be affected  in significantly different ways.

Here we use a ‘niche based’ approach (Leibold 1995) to resource competition to extend the resource-ratio theory of competition into space. We consider two consumers growing on two essential nutrient resources (Tilman 1982) in a spatially heteroegenous metaecosystem of two patches (Gravel et al. 2010a).  We account for internal resource recycling within each patch (Daufresne and Hedin) and ask how the diffusive exchanges of dissolved nutrients, detritus or consumers may alter ceoxistence and ecosystem properties such as nutrient availability, and primary productivity. We find that each type of exchange affects the process in fundamentally distinct ways. Diffusion affect the stoichiometry of the exchanges between patches due to variability in the stoichiometry of each type of material and on the indirect effects they have on resource competition. Detritus exchanges homogenize the productivity of the two patches without affecting any local stoichiometric effects, that exchanges of dissolved nutrients will homogenize both the productivity and stoichiometry of the two patches, and that exchanges of live consumers will have effects that maintain stoichiometry but affect nutrient depletion in patches.  We also show that the joint effects of detritus exchanges can affect coexistence and may generate alternate stable states in local communities.
